---
title: "data.table project"
author: "Eugenia Olari"
date: "2026-03-24"
format: html
---

## 1. Load data adn convert to data.table

```{r}
## load data
dta <- readRDS(file = "data/prec_data.rds")

## load package
library(data.table)

## convert to data.table
dta_t <- as.data.table(x = dta)
print(head(dta_t))
```



## 2. Basic inspection

```{r}
## check classes
class(x = dta)
class(x = dta_t)
```

```{r}
## show first rows
head(x = dta_t)
```

```{r}
## object sizes
object.size(x = dta)
object.size(x = dta_t)
```

```{r}
## column names
names(x = dta_t)
```

```{r}
## dimensions - number of rows and columns
nrow(x = dta_t)
ncol(x = dta_t)
```

```{r}
## column types
str(object = dta_t)
```

```{r}
## check for missing datetimes
sum(is.na(x = dta_t$DT))
```

```{r}
## check QUALITY values
table(dta_t$QUALITY, useNA = "always")
```

```{r}
## check FLAG values
table(dta_t$FLAG, useNA = "always")
```

Based on the inspection above:

- **Identifier variables**: `STATION`, `DT` (datetime), `ELEMENT` — these uniquely identify
each measurement. `STATION` has 431 levels, `ELEMENT` has only one level (`SRA1H`).

- **Categorical variables**: `STATION` and `ELEMENT` are factors. `FLAG` is character,
`QUALITY` is numeric.

- **Variables needing cleaning**: `X` is entirely `NA` and will be dropped. `FLAG` has
136,458 empty strings and 56,382 `"Z"` codes worth checking. `QUALITY` has 128,245 rows
with code `3` and 37,002 with code `4`, indicating lower quality measurements.




## 3. Filtering and summaries

```{r}
## filter VALUE > 0
nz <- dta_t[VALUE > 0, .(STATION, DT, VALUE, ELEMENT, FLAG, QUALITY)]
```

```{r}
## summary by station
nz[, .(mean = mean(x = VALUE),
       sd   = sd(x = VALUE),
       iqr  = IQR(x = VALUE),
       min  = min(x = VALUE),
       max  = max(x = VALUE)),
   by = STATION][order(-mean)]
```

```{r}
## summary by station and year of DT
nz[, .(mean = mean(x = VALUE),
       sd   = sd(x = VALUE),
       iqr  = IQR(x = VALUE),
       min  = min(x = VALUE),
       max  = max(x = VALUE)),
   by = .(STATION, year(x = DT))][order(-mean)] # do the grouped summary
```

The station-level summary covers 431 stations with non-zero precipitation records. 
The highest mean is at station `0-20000-0-11787` (1.04 mm/hr), the lowest at `0-20000-0-11464` (0.56 mm/hr). 
The maximum single-hour value of **64.0 mm** at station `0-203-0-11721` is a clear outlier.

When grouped by year, station `0-20000-0-11787` in **2020** has the highest mean (1.33mm/hr). 
Station `0-203-0-11721` in 2024 has the largest standard deviation (3.50), driven by the 64.0 mm extreme value.




## 4. Duplicate analysis

```{r}
## count exact duplicate rows
sum(duplicated(x = dta_t))
```
The dataset contains **2846 exact duplicate rows**.

```{r}
## show examples of exact duplicates
dta_t[duplicated(x = dta_t) | duplicated(x = dta_t, fromLast = TRUE)][1:10]
```
All duplicates share the timestamp `2018-03-25 01:00:00` — a daylight saving time clock change date, where 01:00 occurs twice. 
Since both rows in each pair have identical values, removing one copy with `unique()` is safe.

```{r}
## define logical duplicate: same station, dt, and element
## count logical duplicates using data.table
dt_dup <- dta_t[, .(n = .N),
                by = .(STATION, DT, ELEMENT)]
```

```{r}
## how many key combinations appear more than once
dt_dup[n > 1, .N]
```

```{r}
## show examples
dt_dup[n > 1][order(-n)][1:10]
```

There are **2922 logical duplicates** — `STATION`, `DT`, `ELEMENT` combinations appearing
more than once. All have `n = 2` and all share the same DST timestamp.



## 5. Cleaning using :=

```{r}
## drop stray X column
dta_t[, X := NULL]
```
The X column contained only NA values across all 25.7 million rows and carries no information


```{r}
## drop rows where DT is NA
dta_t <- dta_t[!is.na(x = DT)]
sum(is.na(x = dta_t$DT))
```
No `NA` values exist in `DT` 

```{r}
## flag suspicious values: negative precipitation is physically impossible
dta_t[, suspicious := VALUE < 0]
dta_t[suspicious == TRUE, .N]
```
No negative `VALUE` entries found 

```{r}
## flag low quality rows
dta_t[, low_quality := QUALITY %in% c(3, 4)]
dta_t[low_quality == TRUE, .N]
```
**165247 rows** have `QUALITY` code `3` or `4` and are flagged as `low_quality`. 
These rows are kept but we will kepp them in mind for cautious use.

```{r}
## remove exact duplicates
dta_t <- unique(x = dta_t)
nrow(x = dta_t)
```
After deduplication the dataset contains **25725335 rows**, down from 25728181 - a reduction of 2846 duplicate rows.


## 6. Renaming and column work

```{r}
## rename DT to DATETIME for clarity
setnames(x = dta_t,
         old = c("DT"),
         new = c("DATETIME"),
         skip_absent = TRUE)

## create year column — useful for temporal grouping
dta_t[, yr := year(x = DATETIME)]

## create row id within each station
dta_t[, row_id := 1:.N,
      by = STATION]

## remove row_id helper column after inspection
dta_t[, row_id := NULL]
names(x = dta_t)
```

`DT` is renamed to `DATETIME` for clarity using `setnames()`. 
A `yr` column is added for fast year-level grouping. 
A `row_id` helper column is created within each station and immediately removed, it was used only to verify row ordering.



## 7.Station summary

```{r}
## station summary
st_sum <- dta_t[, .(
  n_rows = .N,
  mean_val = mean(x = VALUE, na.rm = TRUE),
  sd_val = sd(x = VALUE, na.rm = TRUE),
  min_val = min(x = VALUE, na.rm = TRUE),
  max_val = max(x = VALUE, na.rm = TRUE),
  n_nonzero = sum(VALUE > 0, na.rm = TRUE),
  n_suspicious = sum(suspicious == TRUE, na.rm = TRUE),
  n_flagged = sum(!is.na(x = FLAG) & FLAG != "", na.rm = TRUE),
  n_low_qual = sum(low_quality == TRUE, na.rm = TRUE)
), by = STATION]
```

```{r}
## Q1: most problematic (most flagged + low quality)
st_sum[order(-n_flagged)][1:5]

## Q2: highest non-zero precipitation
st_sum[order(-mean_val)][1:5]

## Q3: most records
st_sum[order(-n_rows)][1:5]

## Q4: most suspicious rows
st_sum[order(-n_suspicious)][1:5]

```

## 8. Element summary

```{r}
## element summary
el_sum <- dta_t[, .(
  n_rows = .N,
  n_stations = uniqueN(x = STATION),
  mean_val = mean(x = VALUE, na.rm = TRUE),
  sd_val = sd(x = VALUE, na.rm = TRUE),
  prop_missing = mean(x = is.na(x = VALUE)),
  n_flagged = sum(!is.na(x = FLAG) & FLAG != "", na.rm = TRUE),
  n_low_quality = sum(low_quality  == TRUE, na.rm = TRUE)
), by = ELEMENT]

## view element summary
el_sum[order(prop_missing)]
```

Since `ELEMENT` has only one level (`SRA1H`), the summary contains a single row.
`SRA1H` covers all 431 stations and 25725335 rows. 
The mean value is 0.081 mm/hr across all records (including zeros), with a proportion of missing values of 0.0014 - very low. There are 192840 flagged rows and 165239 low quality rows.

Since there is only one element type, a cross-element comparison is not possible.
`el_sum` serves here as a data integrity check rather than a comparative analysis.
All quality and reliability questions therefore fall to the station level, which was covered in the previous section.



## 10. Merge task

```{r}
## rebuild nz from cleaned data (non-zero values only)
nz <- dta_t[VALUE > 0]

## create helper table: a regular hourly sequence with VALUE = 0
## this represents the "background" — all hours in the study period
dt <- data.table(VALUE = 0,
                 DATETIME = seq(from = as.POSIXct(x = "2018-01-01 00:00:00"),
                                to = as.POSIXct(x = "2026-01-01 00:00:00"),
                                by = "hour"))

## merge: left join of hourly sequence onto observed non-zero data
## rows in dt that have no matching observation keep VALUE = 0
merged <- merge(x = dt,
                y = nz[, .(DATETIME, VALUE, STATION)],
                by = "DATETIME",
                all.x = TRUE)

## inspect merged result
head(x = merged)
nrow(x = merged)
```

The helper table `dt` contains a complete hourly sequence from 2018 to 2026 **70129 hours** in total. 
Each row has `VALUE = 0`, representing "no recorded precipitation" as a baseline for every hour in the study period.

After the left join, `merged` has **2716679 rows**, which is larger than `dt` because multiple stations can match the same hour.

**What the helper table represents**: a complete time grid covering the full study period. 
Every hour gets a baseline `VALUE = 0`.

**How the merge works**: `nz`, aka non-zero observations, is joined onto `dt` using `DATETIME` as the key.
Hours with a real observation get the actual value filled in via `VALUE.y`. 
Hours with no matching observation keep `VALUE = 0` from `dt` and show `NA` in `VALUE.y` and `STATION`. 
All hours in the grid are preserved because we use `all.x = TRUE`.

**Why this is useful**: it produces a gapless time series, making it easy to detect missing observation periods or compute time-averaged means without gaps distorting the results.



## 11. Reshape task

```{r}
## create a small station-year summary for reshaping
st_yr <- dta_t[VALUE > 0,
               .(mean_val = mean(x = VALUE, na.rm = TRUE)),
               by = .(STATION, yr)]

## reshape to wide: one row per station, one column per year
wide <- dcast(data = st_yr,
              formula = STATION ~ yr,
              value.var = "mean_val")

## inspect wide format
head(x = wide)
```


```{r}
## reshape back to long
long <- melt(data = wide,
             id.vars = "STATION",
             variable.name = "year",
             value.name = "mean_val")

## inspect long format
head(x = long)
```

`dcast()` reshapes the data into wide format — one row per station, one column per year. 
This makes it easy to compare mean precipitation across years side by side.

The output shows stable values across 2018–2024 for most stations, with a slight increase visible in 2024.

`melt()` converts the wide table back to long format, which is better suited for further grouped analysis or plotting.



## 12.  Final conclusions
**Main data problems:**
- `X` column was entirely `NA` — probbaly it was an artifact from data export
- 2846 duplicate rows caused by a DST clock change on `2018-03-25 01:00:00`
- 165247 rows with low quality codes (`3` or `4`)
- 192840 flagged rows across `FLAG` codes `""` and `"Z"`

**What was cleaned:**
- `X` column dropped
- Exact duplicates removed with `unique()`
- `low_quality` and `suspicious` flags added for downstream filtering
- `DT` renamed to `DATETIME`, `yr` column added

**Uncertain decisions:**
- Low quality rows were flagged but kept — whether to exclude them depends
on the analysis
- Empty string `FLAG == ""` rows were kept — unclear if they represent a
logging error or valid unflagged observations

**Most suspicious:**
- Station `0-20000-0-11747` had the most flagged rows - 16741
- Station `0-203-0-11721` had the most extreme single value — 64.0 mm in
one hour, over 20 standard deviations above its mean
- Since `ELEMENT` has only one level, all quality concerns are at the
station level

**Most useful `data.table` operations:**
- `:=` for adding and removing columns in-place
- `by` with `.N` for grouped summaries and duplicate detection
- `dcast()` and `melt()` for reshaping
- `merge()` with `all.x = TRUE` for building a gapless time series
- `setnames()` for safe column renaming
